Automated Formal Synthesis\\ of Safe Digital Controllers for Continuous Plants

### 2019-01-22 automatica/paper.tex
sections

In this work we employ the notion of asymptotic stability within the CEGIS loop,  
as a means for reducing the search space of possible safe controllers, where the definition of a safe controller
is defined in the following section. 
As discussed later, for linear systems a safe controller is necessarily asymptotically stable, 
although the reverse is not true. 
%as an objective for guessing controllers that are later proven sound over safety requirements.  
Qualitatively, (local) asymptotic stability is a property denoting the convergence of the model executions towards an equilibrium point, 
starting from any states in a neighborhood of the point. 
In the case of linear systems %as in~\eqref{eq:closedloopss}, 
considered with a zero reference signal, 
the equilibrium point of interest is the origin (see Figure~\ref{fig:ct} for the portrait of an asymptotically stable execution, converging to the origin).

Of interest for this work, 
it can be shown that a discrete-time LTI system %as \eqref{eq:closedloopss} 
is asymptotically stable if all the roots of its characteristic polynomial (i.e., the
eigenvalues of the closed-loop matrix $A_d - B_d K$) are inside the unity
circle of the complex plane, i.e., if their absolute values are strictly less than
one~\cite{astrom1997computer}. 
Whilst this simple sufficient condition can be either generalized or strengthen to be necessary, 
this is not necessary in the context of this work.    
What is technically key is that in this paper, 
we shall express this asymptotic stability as a specification $\phi_\mathit{stability}$, 
and encode it in terms of a check known as \emph{Jury's criterion}~\cite{fadali}: 
this is an easy algebraic formula to check the entries of matrix $K$, 
so that the closed-loop dynamics are shaped as desired.

@@

A key contribution of this work is that, when synthesizing controllers, it precisely models the finite-precision arithmetic in the digital controller, guaranteeing that these controllers implemented with finite-precision arithmetic are safe. 
Finite-precision arithmetic introduces errors, as shown in Figure~\ref{fig:digitalsystem}. 
Details of how we model this arithmetic are in Section~\ref{sec:numeric_rep}. 
These errors are present in the real world, and we must model them accurately in order to capture the full behaviour of the model:

@@

\item The DAC can potentially introduce further errors. 
However, we assume that the DAC converts numbers to analog signals with the same precision as the controller. 
As a result we can ignore any additional errors introduced by the DAC. 
Again, this assumption is practically reasonable. 
It would, however, be straightforward to account for a DAC or ADC of different precision to the controller in our algorithm, if necessary. 
\end{itemize}

The representations used in the plant model and its arithmetic operations are also carried out at a finite precision, 
thus we must account for the corresponding errors.  
\note{rephrase next:} Specifically, our model of the plant variables $x(t)$ and $u(t)$, which are, in reality, continuous quantities, with finite precision in our model. 
We guarantee that the precision we use to represent the plant variables is at least as precise as the precision used in the digital controller and we use interval arithmetic to bound the errors, as detailed in Section~\ref{sec:numeric_rep}.

@@

In this paper we will use $\mathcal{F}_{\langle I,F \rangle}(x)$ to denote a real number $x$ expressed at a fixed-point precision, 
using $I$ bits to represent the integer part of the number and $F$ bits to represent its decimal part.  
$\mathcal{F}_{\langle I_c,F_c \rangle}(x)$ denotes a real number $x$ represented at the fixed-point precision of the controller, 
and $\mathcal{F}_{\langle I_p,F_p \rangle}(x)$ denotes a real number $x$
represented at the fixed-point precision of the plant model. 
$I_c$ and $F_c$ are determined by the controller.  We pick $I_p$ and $F_p$ for our synthesis
such that $I_p \geq I_c$ and $\allowbreak F_p \geq F_c$, so that our controller 
can be represented accurately in the plant domain.    
Thus any mathematical operations in our modelled digital controller will be in the
range of $\mathcal{F}_{\langle I_c,F_c \rangle}$, and all other calculations
in our model will be carried out in the range of $\mathcal{F}_{\langle
I_p,F_p \rangle}$.

@@

%While in the previous section we have discussed the generic CEGIS framework, 
We are now equipped to apply the general CEGIS framework to the synthesis of digital controllers. 
We start by describing the property that we pass to the synthesizer as the specification $\phi$. 
Namely, as are interested in capturing safety (to be discussed in Section~\ref{ssec:safespecification}) \note{fix?},  
we use a stability specification to narrow the search space of possible controllers, as detailed in Section~\ref{ssec:stabspecification}.

@@

As discussed in Section~\ref{sec:errors}, 
the considered models must account for errors deriving from several sources: 
we must bound the numerical error introduced by the finite-precision representation of the plant (and its operations), 
and precisely model the error introduced by the ADC/DAC conversions, 
as well as the behaviour of the limited-precision arithmetic used by the controller. 
We use interval arithmetic to bound the error introduced by the finite-precision plant, 
and bounded model checking to precisely model the error introduced by the ADC/DAC and the finite-precision controller.

@@

\subsection{Interval arithmetic over errors in numeric representation} 
We use finite-precision arithmetic to model the plant. 
This is an approximation that speeds up each CEGIS iteration, 
however it necessitates a further stage where we verify that the errors introduced by the approximation have not resulted in a controller that is unsafe when executed on a plant expressed over real numbers.  
In this stage, we represent the plant using double-precision floating-point
and we use the Boost interval arithmetic library~\cite{DBLP:journals/tcs/BronnimannMP06} to bound the error in this representation. 
We employ a compositional numeric library to model the fixed point arithmetic for the controller~\cite{CNL} within double-precision floating-point numbers. 
We check the controller is safe starting from each vertex of the set of initial states,  
and show that this is sufficient to prove safety from any state in this set.

@@

Let us first consider the effect of the quantization errors on safety. 
Within the controller, state values are manipulated at low precision,
by means of the vector multiplication $Kx$.
The inputs are thus computed using the following equation: 
%
\begin{align*}
u_{k}&=-(\mathcal{F}_{\langle I_c,F_c \rangle}(K)\cdot\mathcal{F}_{\langle I_c,F_c \rangle}(x_{k})). 
\end{align*}
This induces two types of the errors, 
as detailed above: 
first, the truncation error due to the representation of $x_k$ as $\mathcal{F}_{\langle I_c,F_c
\rangle}(x_{k})$; and second, the rounding error introduced by the multiplication operation. 
Recall this error is modelled precisely by bounded model checking.

@@

\subsubsection{Stability of closed-loop models with fixed-point controller error}
%-------------------------------
\label{sec:stab_FWL}
The validity of Jury's criterion~\cite{fadali} relies on the representation of the closed-loop dynamics $x_{k+1} = (A_d - B_dK) x_k$ at infinite precision.  
When we employ a fixed-point arithmetic digital controller, the operation above can be expressed as follows, where we use $\mathcal{F}_{\langle I_c,F_c \rangle}$ preceding a variable to indicate that variable is converted into the fixed-point precision given by $\mathcal{F}_{\langle I_c,F_c \rangle}$: 
\note{the first of the two equations is not clear}
%
\begin{align*}
x_{k+1} &= A_d \cdot x_{k} -(\mathcal{F}_{\langle I_c,F_c \rangle}(K)\cdot\mathcal{F}_{\langle I_c,F_c \rangle}(x_{k})).  \\
x_{k+1} &= (A_d  - B_dK) \cdot x_k + B_dK\delta, 
%\\ x_{k+1}&=A_d x_{k} + B_du_{k}
\end{align*}
%
where $\delta$ is the maximum error that can be introduced by the digital controller in one step, 
i.e., by reading the states values once and multiplying by $K$ once.  
We derive the closed form expression for $x_n$ recursively, as follows:
%
\small
\begin{align*}
x_{1} &= (A_d  - B_dK)x_0 + B_dK\delta \\
x_{2} %&= (A_d  - B_dK)((A_d  - B_dK)x_0 + B_dK\delta ) + B_dK\delta \\
 &=(A_d  - B_dK)^2x_0 + (A_d  - B_dK)B_dK\delta + B_dK\delta \\
%x_{3}  &=(A_d  - B_dK)^3x_0 + (A_d  - B_dK)^2B_dK\delta + (A_d  - B_dK)B_dK\delta + B_dK\delta\\
x_{n} &= (A_d  - B_dK)^nx_0 + (A_d  - B_dK)^{n-1}B_dK\delta +\\
&\phantom{{}=1} ... + (A_d  - B_dK)^1B_dK \delta + B_dK\delta \\
  &= (A_d - B_dK)^nx_0 + \sum_{i=0}^{i=n-1}(A_d - B_dK)^iB_dk\delta. 
\end{align*}
\normalsize
%The definition of asymptotic stability is that the system converges to a
%reference signal, in this case we use no reference signal so 
Recall that a closed-loop asymptotically stable system will converge to the origin.  
We know that the original system with an infinite-precision controller is stable, 
because we have synthesized it to meet Jury's criterion.  
Hence, $(A_d - B_dK)^n x_0$  must converge to zero as $n \uparrow \infty$. 
%T%o show that a fixed-point implementation of the same controller will
%converge to a finite value, we need to show that the infinite sum converges. 
Furthermore, the power series of a square matrix $T$ converges~\cite{horn1990matrix} iff the eigenvalues of the matrix are less
than~1, and the limit results in 
%
%\begin{align*}
$\sum_{i=0}^{\infty}T^i  = (I - T)^{-1}$, 
%\end{align*}
%
where $I$ is the identity matrix. 
Thus, the closed-loop model converges to the value 
%
\begin{align*}
0 + (I - A_d + B_dK)^{-1}B_dk\delta \,. 
\end{align*}
%
As a result, if the value $(I - A_d + B_dK)^{-1}B_dk\delta$ is within the safe set of states given by the safety specification
then the synthesized fixed-point controller results in a safe
closed-loop model.  The convergence to a finite value, however, will not
make it asymptomatically stable.
However,
since in this paper we require stability only as a precursor to safety, it
is sufficient to check that the (perturbed, noisy) model converges to a
neighborhood of the equilibrium within the safe set.

>>>>

In this work we employ the notion of asymptotic stability within the CEGIS loop,  
as a means for reducing the search space of possible safe controllers, where the definition of a safe controller
is defined in the following section. 
As discussed later, specifically for linear models a safe controller is necessarily asymptotically stable, 
although the reverse is not true. 
%as an objective for guessing controllers that are later proven sound over safety requirements.  
Qualitatively, (local) asymptotic stability is a property denoting the convergence of the model executions towards an equilibrium point, 
starting from any states in a neighborhood of the point. 
In the case of linear systems %as in~\eqref{eq:closedloopss}, 
considered with a zero reference signal, 
the equilibrium point of interest is the origin (see Figure~\ref{fig:ct} for the portrait of an asymptotically stable execution, converging to the origin).

Of interest for this work, 
it can be shown that a discrete-time LTI model %as \eqref{eq:closedloopss} 
is asymptotically stable if all the roots of its characteristic polynomial (i.e., the
eigenvalues of the closed-loop matrix $A_d - B_d K$) are inside the unity
circle of the complex plane, i.e., if their absolute values are strictly less than
one~\cite{astrom1997computer}. 
Whilst this simple sufficient condition can be either generalized or strengthen to be necessary, 
this is not necessary in the context of this work.    
What is technically key is that in this paper, 
we shall express this asymptotic stability as a specification $\phi_\mathit{stability}$, 
and encode it in terms of a check known as \emph{Jury's criterion}~\cite{fadali}: 
this is an easy algebraic formula to check the entries of matrix $K$, 
so that the closed-loop dynamics are shaped as desired.

@@

A key contribution of this work is that, when synthesizing controllers, it precisely models the finite-precision arithmetic in the digital controller, guaranteeing that these controllers implemented with finite-precision arithmetic are safe. 
Finite-precision arithmetic introduces errors, as shown in Figure~\ref{fig:digitalsystem}. 
Details of how we model this arithmetic are in Section~\ref{sec:numeric_rep}. 
\note{[rephrase?] These errors are present in the real world}, and we must model them accurately in order to capture the full behaviour of the model:

@@

\item The DAC can potentially introduce further errors. 
However, we assume that the DAC converts numbers to analog signals with the same precision as the controller. 
Again, this assumption is practically reasonable. 
As a result we can ignore any additional errors introduced by the DAC. 
It would, however, be straightforward to account for a DAC or ADC of different precision to the controller in our algorithm, if necessary. 
\end{itemize}

\note{[shouldn't this be a bullet point too?]}
The representations used in the plant model and its arithmetic operations are also carried out at a finite precision, 
thus we must account for the corresponding errors.  
\note{rephrase next - main clause has no verb:} Specifically, our model of the plant variables $x(t)$ and $u(t)$, which are, in reality, continuous quantities, with finite precision in our model. 
We \note{guarantee [?]} that the precision we use to represent the plant variables is at least as precise as the precision used in the digital controller and we use interval arithmetic to bound the errors, as detailed in Section~\ref{sec:numeric_rep}.

@@

In this paper we will use $\mathcal{F}_{\langle I,F \rangle}(x)$ to denote a real number $x$ expressed at a fixed-point precision, 
using $I$ bits to represent the integer part of the number and $F$ bits to represent its decimal part.  
In particular, $\mathcal{F}_{\langle I_c,F_c \rangle}(x)$ denotes a real number $x$ represented at the fixed-point precision of the controller, 
and $\mathcal{F}_{\langle I_p,F_p \rangle}(x)$ denotes a real number $x$
represented at the fixed-point precision of the plant model. 
$I_c$ and $F_c$ are determined by the controller.  We pick $I_p$ and $F_p$ for our synthesis
such that $I_p \geq I_c$ and $\allowbreak F_p \geq F_c$, so that our controller 
can be represented accurately in the plant domain.    
Thus any mathematical operations in our modelled digital controller will be in the
range of $\mathcal{F}_{\langle I_c,F_c \rangle}$, and all other calculations
in our model will be carried out in the range of $\mathcal{F}_{\langle
I_p,F_p \rangle}$.

@@

%While in the previous section we have discussed the generic CEGIS framework, 
We are now equipped to apply the general CEGIS framework to the synthesis of digital controllers. 
We start by describing the property that we pass to the synthesizer as the specification $\phi$. 
Namely, as we are interested in capturing safety (to be discussed in Section~\ref{ssec:safespecification}\note{fix?}),  
we use a stability specification to narrow the search space of possible controllers, as detailed in Section~\ref{ssec:stabspecification}.

@@

As discussed in Section~\ref{sec:errors}, 
the considered models must account for errors deriving from several sources: 
we must bound the numerical error introduced by the finite-precision representation of the plant (and its operations), 
and precisely model the error introduced by the ADC/DAC conversions, 
as well as the behaviour of the limited-precision arithmetic used by the controller.

We employ interval arithmetic to bound the error introduced by the finite-precision plant, 
and bounded model checking to precisely model the error introduced by the ADC/DAC and the finite-precision controller.

@@

\subsection{Interval arithmetic over errors in numeric representation} 
We use finite-precision arithmetic to model the plant. 
This is an approximation that speeds up each CEGIS iteration, 
however it necessitates a further stage where we verify that the errors introduced by the approximation have not resulted in a controller that is unsafe when executed on a plant expressed over real numbers.  
In this stage, we represent the plant using double-precision floating-point
and we use the Boost interval arithmetic library~\cite{DBLP:journals/tcs/BronnimannMP06} to bound the error in this representation. 
We employ a compositional numeric library to model the fixed point arithmetic for the controller~\cite{CNL} within double-precision floating-point numbers. 
We check that the controller is safe starting from each vertex of the set of initial states,  
and show that this is sufficient to prove safety from any state in this set.

@@

Let us first consider the effect of the quantization errors on safety. 
Within the controller, state values are manipulated at low precision,
by means of the vector multiplication $Kx$.
The inputs are thus computed using the following equation: 
%
\begin{align*}
u_{k}&=-(\mathcal{F}_{\langle I_c,F_c \rangle}(K)\cdot\mathcal{F}_{\langle I_c,F_c \rangle}(x_{k})). 
\end{align*}
This induces two types of errors, 
as detailed above: 
first, the truncation error due to the representation of $x_k$ as $\mathcal{F}_{\langle I_c,F_c
\rangle}(x_{k})$; and second, the rounding error introduced by the multiplication operation. 
Recall this error is modelled precisely by bounded model checking.

@@

\subsubsection{Stability of closed-loop models with fixed-point controller error}
%-------------------------------
\label{sec:stab_FWL}
The validity of Jury's criterion~\cite{fadali} relies on the representation of the closed-loop dynamics $x_{k+1} = (A_d - B_dK) x_k$ at infinite precision.  
When we employ a fixed-point arithmetic digital controller, the operation above can be expressed as follows, where we use $\mathcal{F}_{\langle I_c,F_c \rangle}$ preceding a variable to indicate that variable is converted into the fixed-point precision given by $\mathcal{F}_{\langle I_c,F_c \rangle}$: 
\note{the first of the two equations is not clear - don't you miss a B matrix?}
%
\begin{align*}
x_{k+1} &= A_d \cdot x_{k} -(\mathcal{F}_{\langle I_c,F_c \rangle}(K)\cdot\mathcal{F}_{\langle I_c,F_c \rangle}(x_{k})).  \\
x_{k+1} &= (A_d  - B_dK) \cdot x_k + B_dK\delta, 
%\\ x_{k+1}&=A_d x_{k} + B_du_{k}
\end{align*}
%
where $\delta$ is the maximum error that can be introduced by the digital controller in one step, 
i.e., by reading the states values once and multiplying by $K$ once.  
We derive the closed form expression for $x_n$ recursively, as follows:
%
\small
\begin{align*}
x_{1} &= (A_d  - B_dK)x_0 + B_dK\delta \\
x_{2} %&= (A_d  - B_dK)((A_d  - B_dK)x_0 + B_dK\delta ) + B_dK\delta \\
 &=(A_d  - B_dK)^2x_0 + (A_d  - B_dK)B_dK\delta + B_dK\delta \\
%x_{3}  &=(A_d  - B_dK)^3x_0 + (A_d  - B_dK)^2B_dK\delta + (A_d  - B_dK)B_dK\delta + B_dK\delta\\
x_{n} &= (A_d  - B_dK)^nx_0 + (A_d  - B_dK)^{n-1}B_dK\delta +\\
&\phantom{{}=1} ... + (A_d  - B_dK)^1B_dK \delta + B_dK\delta \\
  &= (A_d - B_dK)^nx_0 + \sum_{i=0}^{i=n-1}(A_d - B_dK)^iB_dk\delta. 
\end{align*}
\normalsize
%The definition of asymptotic stability is that the system converges to a
%reference signal, in this case we use no reference signal so 
Recall that a closed-loop asymptotically stable system will converge to the origin.  
We know that the original system with an infinite-precision controller is stable, 
because we have synthesized it to meet Jury's criterion.  
Hence, $(A_d - B_dK)^n x_0$  must converge to zero as $n \uparrow \infty$. 
%T%o show that a fixed-point implementation of the same controller will
%converge to a finite value, we need to show that the infinite sum converges. 
Furthermore, the power series of a square matrix $T$ converges~\cite{horn1990matrix} iff the eigenvalues of the matrix are less
than~1, and the limit results in 
%
%\begin{align*}
$\sum_{i=0}^{\infty}T^i  = (I - T)^{-1}$, 
%\end{align*}
%
where $I$ is the identity matrix. 
Thus, the closed-loop model converges to the value 
%
\begin{align*}
0 + (I - A_d + B_dK)^{-1}B_dk\delta \,. 
\end{align*}
%
As a result, if the value $(I - A_d + B_dK)^{-1}B_dk\delta$ is within the safe set of states given by the safety specification
then the synthesized fixed-point controller results in a safe
closed-loop model.  The convergence to a finite value, however, will not
make it asymptomatically stable.
However,
since in this paper we require stability only as a precursor to safety, it
is sufficient to check that the perturbed model converges to a
neighborhood of the equilibrium within the safe set.
